% ============================================================
%  Chapter 1 --- Introduction
% ============================================================
\chapter{Introduction}
\label{ch:introduction}

\section{Project Overview}

\shiftdb{} is a fully functional relational database management system built entirely from scratch.  The project spans every layer of a modern DBMS --- from raw byte-level page storage through SQL parsing and query execution to a premium React frontend --- making it an ideal platform for understanding database internals at a depth suitable for both academic study and technical interviews.

\begin{conceptbox}[Why Build a DBMS from Scratch?]
Building a DBMS from the ground up forces you to confront the same engineering trade-offs that the designers of PostgreSQL, MySQL, and SQLite face every day:
\begin{itemize}[nosep]
  \item How do you lay out data on disk so that reads are fast?
  \item How do you parse an expressive query language efficiently?
  \item How do you maintain data integrity under concurrent access?
  \item How do you recover gracefully from crashes?
\end{itemize}
Answering these questions with working code is the deepest form of learning.
\end{conceptbox}

\section{Motivation}

\subsection{Bridging Theory and Practice}
University courses introduce B-trees, buffer pools, and transaction isolation levels in the abstract.  \shiftdb{} makes every concept concrete: the B-tree lives in \code{indexing} logic, the buffer pool is a real LRU cache backed by \code{buffer\_pool.h}, and transactions are managed with actual locks.  Each chapter of this document maps one theoretical topic to the corresponding implementation in the codebase.

\subsection{Interview Preparation}
Database internals questions appear frequently in systems-design and backend-engineering interviews.  This guide pairs each technical deep-dive with curated interview-style questions (see each chapter's \emph{Interview Corner} and \cref{ch:interview-prep} for a comprehensive Q\&A bank).

\section{Technology Stack}

\begin{table}[H]
\centering
\caption{Technology stack of \shiftdb{}.}
\label{tab:tech-stack}
\begin{tabular}{l p{12cm}}
\toprule
\textbf{Component} & \textbf{Technology} \\
\midrule
Storage Engine        & Custom C++20 (slotted pages, buffer pool with LRU eviction) \\
SQL Parser            & Hand-written recursive-descent parser in C++ \\
Node.js Backend       & Express + better-sqlite3 (alternative deployment) \\
Frontend Framework    & React 18 + TypeScript \\
Build Tool            & Vite 8 \\
SQL Editor            & Monaco Editor (VS Code engine) \\
Diagrams              & ReactFlow (xyflow) \\
Charts                & Recharts \\
Animations            & Framer Motion \\
State Management      & Zustand \\
Styling               & Tailwind CSS v4 + CSS custom properties \\
Icons                 & Lucide React \\
\bottomrule
\end{tabular}
\end{table}

\section{Feature Summary}

\subsection{C++ Backend}
\begin{itemize}[nosep]
  \item \textbf{Page-based storage engine} with slotted-page format and buffer pool (LRU).
  \item \textbf{Full SQL parser}: recursive descent supporting the complete SQL grammar.
  \item \textbf{Query executor}: \code{CREATE}, \code{INSERT}, \code{SELECT}, \code{UPDATE}, \code{DELETE}, \code{DROP}, \code{SHOW}, \code{DESCRIBE}, \code{EXPLAIN}.
  \item \textbf{WHERE clause engine}: \code{=}, \code{!=}, \code{>}, \code{<}, \code{>=}, \code{<=}, \code{AND}, \code{OR}, \code{LIKE}, \code{IN}, \code{BETWEEN}, \code{IS NULL}.
  \item \textbf{JOIN operations}: \code{INNER}, \code{LEFT}, \code{RIGHT}, \code{FULL}, \code{CROSS}.
  \item \textbf{Aggregates}: \code{COUNT}, \code{SUM}, \code{AVG}, \code{MIN}, \code{MAX} with \code{GROUP BY}/\code{HAVING}.
  \item \textbf{Result-set control}: \code{ORDER BY}, \code{LIMIT}, \code{DISTINCT}.
  \item \textbf{Constraints}: \code{PRIMARY KEY}, \code{UNIQUE}, \code{NOT NULL}, \code{DEFAULT}, \code{AUTO\_INCREMENT}.
  \item \textbf{Catalog persistence}: schema in \code{catalog.json}; data in \code{.nxdb} binary files.
  \item \textbf{Built-in HTTP server}: raw-socket server with zero external dependencies.
  \item \textbf{REST API}: complete CRUD, schema introspection, statistics, export/import.
\end{itemize}

\subsection{React Frontend}
\begin{itemize}[nosep]
  \item \textbf{Monaco SQL Editor} with schema-aware autocomplete.
  \item \textbf{Interactive Data Grid} with sorting, pagination, CSV/JSON export.
  \item \textbf{Schema Explorer} with searchable tree view.
  \item \textbf{Interactive ER Diagram} (ReactFlow) with auto-detected relationships.
  \item \textbf{Visual Query Builder}: drag-and-drop SQL construction.
  \item \textbf{Query Plan Visualiser}: colour-coded tree view of \code{EXPLAIN} output.
  \item \textbf{Performance Monitor}: real-time charts for query timing and buffer-pool hit rate.
  \item \textbf{Query History}: searchable, click-to-replay, with execution timing.
  \item \textbf{Command Palette} (\texttt{Ctrl+K}): fuzzy search across all actions.
  \item \textbf{Dark/Light Theme} with glassmorphism effects.
\end{itemize}

\section{How to Read This Document}

Each chapter follows a consistent structure:
\begin{enumerate}[nosep]
  \item \textbf{Theory} --- the textbook concepts explained with intuition and diagrams.
  \item \textbf{Implementation} --- how \shiftdb{} realises those concepts in code, with annotated listings.
  \item \textbf{Deep Dive} --- subtleties, edge cases, and design trade-offs.
  \item \textbf{Interview Corner} --- curated questions and model answers.
\end{enumerate}

\noindent Chapters are self-contained; feel free to jump to whichever topic you need.  Cross-references (\cref{ch:storage-engine}, \cref{ch:buffer-management}, etc.) link related discussions.

\begin{interviewbox}[Quick Self-Test]
\textbf{Q:} What are the five core components every relational DBMS must have?\\[4pt]
\textbf{A:} (1) Storage manager, (2) buffer manager, (3) query parser/optimiser, (4) execution engine, (5) transaction/concurrency manager.  \shiftdb{} implements all five --- each has its own chapter.
\end{interviewbox}
